On The Subspace Learning for Network Attack Detection

### 2020-04-28 docs/thesis/chapters/4_m_rpca/m_rpca.tex
updating thesis latex

\chapter{Moment Distances from Robust Subspace for Network Anomaly Detection}
\label{ch:4_m_rpca}

\begin{quotation}[]{Paulo Freire}
No one knows it all. No one is ignorant of everything. We all know something. We are all ignorant of something.
\end{quotation}

Some widely adopted algorithms for anomaly detection assume a Gaussian or symmetric distributed data, however this assumption may not be observed in some real world problems, such as the case of network traffic analysis, where network traffic features are usually more characterized by skewed and heavy-tailed distributions \cite{lakhina2005mining,benson2010network, leon2017probability}.

@@

These findings show that the skewness and heavy-tailed distributions may be important for network traffic analysis, and can motivate researches to evaluate the impact of skewed data into algorithms that rely on Gaussian distributed data. These findings can also highlight opportunities to  exploit the skewness and heavy-tailed distributions to obtain accurate classifiers for network anomaly detection.

An outlyingness-approach based on a robust estimator of skewness, combined with robust estimators of location and scale, can be able to flag the outlying measurements. When the same methodology would be used with non-robust estimators of location, scale and skewness, the outlyingness-values would be affected by the outliers such that the outlying group could be masked \cite{hubert2009robustskewed}.

Network anomaly detection problems are usually characterized by imbalanced data \cite{Phua2004minority,he2008learning,benson2010network}. However, learning algorithms for imbalanced data has been a challenging research topic, considering that the fundamental issue with the imbalanced learning problem is the ability of imbalanced data to significantly compromise the performance of most standard learning algorithms \cite{he2008learning}. Therefore, learning methods for imbalanced and skewed data have been attracted attention of researchers \cite{Phua2004minority,hubert2009robustskewed}.

We believe that the skewness of anomalous and normal traffic can highlight features for improving anomaly detection in imbalanced data. We also believe that the distance between robust estimates of normal traffic and contaminated observations can highlight discrepancies and be used for network attack detection.

Therefore, we propose the m-RPCA, which is an approach based on distances between moments computed from a robust subspace learned by Robust Principal Component Analysis (RPCA) and contaminated observations, in order to detect anomalies from skewed data and network traffic. The proposed approach relies on a robust subspace computed from supposed normal observations, for estimating the moments to be used for distance analysis. The anomaly detection from contaminated observations evaluate the Mahalanobis distance between the robust moments and new contaminated observations, in a semi-supervised fashion, without the computational cost of new robust subspace learning for new observations. The m-RPCA can also be computed as an unsupervised algorithm, with subspace learning from the same contaminated data that is the target of the anomaly detection analysis.

We evaluate the accuracy of the m-RPCA for anomaly detection on simulated data sets, with skewed and heavy-tailed distributions, and for the CTU-13 data set \cite{garcia2014empirical}, which is a large data set of normal, background and botnet traffic that has been adopted to deal with the lack of up-to-date real-world data sets for anomaly detection systems \cite{osanaiye2016distributed}. The Experimental evaluation compares our proposal to widely adopted algorithms for anomaly detection based on clustering and statistical approaches. We also compare the results to ROBPCA \cite{hubert2005robpca}, which is a method that also relies on robust estimates with adjusted outlyingness based on robust skewness.

The main contribution of this work is the proposal of a novel semi-supervised and unsupervised method for anomaly detection in skewed and imbalanced data, with results of experimental evaluation on simulated and real data sets.

This Chapter is organized as follows. In Section \ref{sec:4_relatedworks}, it is conducted a literature review of network anomaly detection, botnet detection, and imbalanced learning. Section \ref{sec:4_datamodel} presents the data model and the evaluated data set. Section \ref{sec:4_proposal} describes the proposed approach for network attack detection. Section \ref{sec:4_experiments} discusses the experimental validation, Section \ref{sec:4_simulated_result} presents the results of the simulated scenarios and the Section \ref{sec:4_ctu13_result} describes the results for anomaly detection on CTU-13 data set. Finally, Section \ref{sec:4_conclusion} draws the conclusions and the suggestions for future work.

\section{Related Works}
\label{sec:4_relatedworks}

@@

@@

Khattak \emph{et al.} \cite{khattak2015botflex} proposed BotFlex, which is a network-based tool for botnet detection, composed by a Complex Event Processing (CEP) engine and on a correlation framework that continuously receives events and correlates them according to rules. BotFlex's results are compared to BotHunter \cite{gu2007bothunter}, but the evaluation relies in a own and not public data set.

@@

Considering the lack of available labeled data sets, Garcia \emph{et al.} \cite{garcia2014empirical} proposes the CTU-13 data set, which is composed by attack, normal and background labeled data, in an imbalanced distribution like in a real networks. The authors recommend scenarios for training and testing in order to avoid the use of traffic from a botnet family for training and testing, aiming to ensure that the evaluated methods can generalize and detect new behaviors. Taking into account the adoption of unsupervised or semi-supervised approaches for anomaly detection, adopting the training and testing approach proposed by Garcia \emph{et al.}, some botnet malwares wouldn't be tested, since in the author's proposal some botnets are present only for training.

@@

% Anomaly detection by subspace learning
Traditional PCA-based anomaly detection models are not suitable for anomaly interpretation, as they judge whether a data instance is an anomaly or not based on the length of its projection on the abnormal subspace spanned by the less significant principal components, and there is no direct mapping between PCA’s dimensionality-reduced subspace and the original feature space \cite{ringberg2007sensitivity}. However, to overcome the above mentioned limitations, some approaches based on PCA have been proposed for network anomaly detection. Callegari \emph{et al.} \cite{callegari2011novel} proposed a PCA-based method for identifying the network traffic flows responsible for an anomaly detected at the aggregate level, by means of a separation of normal and anomalies according to principal components (normal) and remaining (anomalies). Lee \emph{et al.} \cite{Lee2013} presented OverSampling PCA (osPCA), which allows one to determine the anomaly of the target instance according to the variation of the resulting dominant eigenvector obtained by similarity analysis and over sampling. Vieira \emph{et al.} \cite{vieira2017model} proposed a framework that applies Model Order Selection (MOS) for detection of time frames under attack and uses similarity analysis to extract details and detect the time and ports under attack.

% Anomaly detection by robust subspace learning
The problem of PCA or subspace learning for outlier corrupted data is called Robust Principal Component Analysis (RPCA) or robust subspace learning \cite{candes2011robust, vaswani2018robust}. RPCA aims to be resilient to outliers by means of a robust subspace learning \cite{vaswani2018robust} for outlier corrupted data, decomposing a given data matrix $\pmb{X}$ into the sum of a low rank matrix $\pmb{L}$, whose column subspace gives the principal components, and a sparse matrix $\pmb{S}$, which refers to outliers’ matrix. This definition is also referred to as the sparse + low rank ($\pmb{S + LR}$) formulation \cite{vaswani2018robust}. We refer to \cite{lerman2018overview} and \cite{vaswani2018robust} for more details regarding robust subspace learning.

RPCA is been mainly applied to computer vision, in problems of robust subspace tracking and robust subspace recovery. However, RPCA has also been adopted for general outlier detection \cite{hubert2005robpca,hubert2009robustskewed,cherapanamjeri2017thresholding,zhou2017anomaly,NetflixSurus} and for anomaly detection on network traffic \cite{pascoal2012robust}. ROBPCA \cite{hubert2005robpca} intends to identify outliers using PCA from robust estimates of mean and covariance matrix, to reduce the data dimensions and plotting the orthogonal distances versus the robust score distances, to flag an outlier map. However, ROBPCA flags many points as outlying when the original data is skewed. Therefore, Hubert \emph{et al.} \cite{hubert2009robustskewed} proposed ROBPCA-AO, which improves ROBPCA for problems with skewed data, by means of an adjusted outlyingness based on robust skewness. Hubert \emph{et al.} \cite{hubert2009robustskewed} evaluated ROBPCA-AO for real and simulated data sets, but it is not clear if this method was evaluated for very skewed and imbalanced data sets, such as in the network attack detection problem.

Robust subspace learning has receiving a growing attention of researchers aiming the development of network anomaly detection systems \cite{rousseeuw1984mcd, rousseeuw1999fastmcd, hubert2005robpca,hubert2009robustskewed, pascoal2012robust, zhou2017anomaly}, considering outlier-robust methods and sparse-corruption methods \cite{lerman2018overview}. Pascoal \emph{et al.} \cite{pascoal2012robust} proposed an approach based on a robust mutual information estimator for feature selection and based on RPCA for outlier detection in internet traffic. The anomaly detection proposed by Pascoal \emph{et al.} is an unsupervised approach that estimate the first $k$ robust principal components, calculate the score and the orthogonal distances, calculate the thresholds and classify new observations accordingly. Zhou and Paffenroth \cite{zhou2017anomaly} proposed the Robust Deep Autoencoders (RDA), which the central idea is that a RDA inherits the non-linear representation capabilities of autoencoders combined with the anomaly detection capabilities of RPCA. Considering that outliers and noise may reduce the quality of representations discovered by deep autoencoders, the proposed model isolates noise and outliers in the input by means of a RPCA approach, and the autoencoder is trained after this isolation. RDA was evaluated by the authors for the MNIST data set.

@@

Mahalanobis Distance (MD) is a generalized distance which is useful for determining the similarity between an unknown sample and a collection of known samples, by considering the covariance between the variables and their mean values. The MD is a measure of the distance between a vector $\pmb{x}$ and a distribution $\pmb{X}$, introduced by P. C. Mahalanobis in 1936 \cite{mahalanobis1936md}. It is a multi-dimensional generalization for measuring how many standard deviations away $\pmb{x}$ is from the mean $\bar{\pmb{x}}$ and covariance $\bar{\pmb{\Sigma}}$ of $\pmb{X}$. MD has been used for distance based anomaly detection with robust estimates in many areas, assuming that the Mahalanobis Distance between robust estimates and contaminated observations can reveal anomalies.

We propose an approach based on the distances between moments computed from learned robust subspace and contaminated observations, for anomaly detection on skewed and imbalanced data sets, and evaluate our approach for network attack detection. The proposed approach relies on learning robust subspace from supposed normal traffic for estimating the moments (mean, skewness and kurtosis). Thus, the anomaly detection for contaminated observations should evaluate the Mahalanobis Distance between the robust moments and contaminated observations, in a semi-supervised fashion and without the computational cost of new robust subspace learning for new observations, or in a unsupervised approach, where the robust subspace is learned from the contaminated data and the distance analysis is between the robust moments and the contaminated observations.

\section{Data Model}
\label{sec:4_datamodel}

In this paper, scalars are denoted by italic letters (\emph{a, b, A, B, $\alpha$, $\beta$}), vectors by lowercase bold letters ($\pmb{a}, \pmb{b}$), matrices by uppercase bold letters ($\pmb{A}$, $\pmb{B}$), and $a_i,_j$ denotes the (\emph{i, j}) elements of the matrix $\pmb{A}$. The superscripts \textsuperscript{T} and \textsuperscript{-1} are used for matrix transposition and matrix inversion, respectively. The Frobenius norm is denoted as $\left\| \mathord{\cdot} \right\|_F$, while $\left\| \mathord{\cdot} \right\|_*$ denotes the nuclear norm of a matrix and $\left\| \mathord{\cdot} \right\|_1$ means the sum of the absolute values of matrix entries. We also define the operator $\langle \mathord{\cdot} \rangle$, that is the standard trace inner product, and $[ \mathord{\cdot} ]^s$, which denotes the $n$ largest values of a vector.

This section also presents a description of the simulated data model in Subsection \ref{sec:4_simulation} and in Subsection \ref{sec:4_CTU-13} we describe the data model for the CTU-13 data set.

\subsection{The Simulated data set}
\label{sec:4_simulation}

In order to analyse the null hypotheses $H_{1rob.sub}$ and $H_{0rob.sub}$, which evaluate if the robust subspace learning can improves the anomaly detection from imbalanced and skewed data, we create two simulated data sets characterized by skewed and heavy tailed distributions, and create a simulated Gaussian distributed data set, in order to also analyse the detection rate for not skewed and not heavy tailed distributions.

We selected Pareto (with scale $a=3$, mode $m=1$ and $\mu=1$) and Lognormal (with mean $\mu=0$ and standard deviation $\sigma=1$) distributions, denoted respectively as $\pmb{Y}_p$, while $\pmb{Y}_l$, to simulate normal events and to evaluate the anomaly detection on skewed data performed by our proposals in comparison to widely adopted algorithms for outlier detection, considering that these distributions are skewed and heavy-tailed, and have been adopted to characterize network traffic of the Internet and data centers \cite{benson2010network,leon2017probability}. We also adopt the Gaussian distributed data set $\pmb{Y}_g$ to simulate normal events with mean $\mu=0$ and unitary standard deviation $\sigma=1$.

@@

An uniform noise between $-6$ and $6$ is used to add anomalies into the Gaussian distributed data set, and $\pmb{Y}_g^c$ denotes the contaminated Gaussian data contaminated by an uniform distribution. The figure \ref{fig:4.04} shows an example of the Gaussian distribution of normal observations and the uniform distribution to simulate the addition of anomalies.

@@

\subsection{The CTU-13 data set}
\label{sec:4_CTU-13}

The CTU-13 \cite{garcia2014empirical} is a data set of botnet traffic that was captured in the Czech Technical University, by means of a testbed and malware execution in a real network. The CTU-13 data set contains 13 scenarios with network flows of botnet malwares, that are: neris, rbot, virut, menti, sogou, nsys.ay and murlo. The botnet traffic is also classified as attack or command and control (C\&C), while the legitimate flows can also be classified as normal or background. It is important to note that the CTU-13 data set can also be formulated as a superposition of signal, artefact and noise, which refer to background, normal a botnet traffic, respectively.

@@

\begin{itemize}
	\item \pmb{Attacks:} Click Fraud (CF), Port Scan (PS), Compiled and Controlled by Authors (US), SPAM and DDOS;
	\item \pmb{C\&C:} IRC, P2P and HTTP.
\end{itemize}

We refer to Garcia \cite{garcia2014identifying} and Garcia \emph{et al.} \cite{garcia2014empirical} for a detailed description of the performed attacks and C\&C flows, as well as for more information about the topology of the adopted testbed, rules for classifying normal flows, and an analysis of behaviors or patterns of the malware's traffic.

For all the scenarios, the authors of the CTU-13 data set convert the captured pcap files to NetFlows and release the processed flows. The data set provides ground-truth labels for flows as follows: flows from or to the infected machines are labeled as “botnet”; flows from or to well known and controlled machines are labeled as “normal”; all other flows are labeled as “background.”

Table \ref{tab:4.01} presents an overview grouped by scenario, according to the column ID, and shows the malwares used for botnet attacks, the types of attacks or C\&C, the total number of flows, the number of malicious flows which includes flows of C\&C and attacks, and finally shows the number of normal flows.

>>>>

\chapter{Robust Framework for Detection of Network Attacks in Imbalanced Traffic}
\label{ch:4_m_rpca}

According to \cite{distilnetworks2019}, 37.9\% of all Internet traffic of 2018 was not from human activities, but from bots, that can be classified as good bots, that perform legitimate operations, and bad bots, which perform malicious activities, such as DDoS attacks, probe attack or frauds. In 2018, bad bots accounted for 20.4\% of all website traffic.

According to \cite{BISSELL2019} the proportion of cost on discovery activities for cybersecurity has increased steadily since 2015. Forensic cyber and user behavior analytics also present an opportunity for cost savings — US\$1.72 million — with discovery and investigation activities. However, only 32\% of organizations have deployed these technologies enterprise-wide. Accenture consulting argues that 89\% of survey respondents believe breakthrough technologies, like machine learning and user behavior analytics, are essential to securing the future of their organizations \cite{Accenture2018}.

To face the adversarial model, network attacks and counter measures of attackers to avoid detection, it is possible to adopt unsupervised or semi-supervised approaches for network anomaly detection, by means of behavioral analysis, where it is not necessary known anomalies for training models \cite{moustafa2019holistic}.

Anomalies in the context of network traffic can be hard to identify and separate from legitimate data due to the rare occurrences of anomalies in comparison to legitimate events. Therefore, anomaly detection algorithms have to be highly discriminating, robust to corruption and able to deal with the imbalanced data problem \cite{he2008learning}. Note that data corruption refers to outliers that can be part of the data but not be considered malicious. The imbalanced data problem corresponds to data sets exhibiting significant imbalances of classes or rare events of some classes \cite{he2009learning}, which can be legitimate or malicious classes in network anomaly detection problems.

Traditionally adopted algorithms for anomaly detection assume a Gaussian or symmetric distributed data \cite{lakhina2005mining}, however this assumption may not be observed in some real world problems \cite{benson2010network}, such as the case of network traffic analysis, where network traffic features are usually more characterized by skewed and heavy-tailed distributions \cite{leon2017probability}.

@@

The findings presented by Benson \emph{et al.} \cite{benson2010network} and Leon-Garcia \cite{leon2017probability} show that the skewness and heavy-tailed distributions may be important for network traffic analysis, and can motivate researches to evaluate the impact of skewed data into algorithms that rely on Gaussian distributed data for network anomaly detection. Additionally, the fitting of network traffic to skewed and heavy-tailed distributions can indicate opportunities to exploit properties and characteristics of the skewness and heavy-tailed distributions to obtain improved classifiers for network anomaly detection.

Network anomaly detection problems are usually characterized by imbalanced data \cite{Phua2004minority}. However, learning algorithms for imbalanced data has been a challenging research topic, considering that the imbalanced data can compromise the performance of most standard learning algorithms, creating bias or unfair weight to learn from the majority class and reducing detection capacity of anomalies that are characterized by the minority class \cite{he2008learning}. Therefore, learning methods for imbalanced and skewed data have attracted attention of researchers \cite{hubert2009robustskewed}.

Robust subspace learning has been attracting a growing attention of researchers aiming the development of network attack detection systems \cite{pascoal2012robust, zhou2017anomaly} that rely on behavioral analysis. An outlyingness-approach based on a robust estimator of skewness, combined with robust estimators of location and scale, can be able to flag the outlying measurements \cite{hubert2009robustskewed}. According to Hubert \emph{et al.} \cite{hubert2009robustskewed}, when the same methodology would be used with non-robust estimators of location, scale and skewness, the outlyingness-values would be affected by the outliers such that the outlying group could be masked.

Considering that the skewness of anomalous and legitimate traffic can highlight features for improving anomaly detection and network attack detection in imbalanced data, and considering that the distance between robust estimates can be used for network attack detection, we propose the Moment-based Robust Principal Component Analysis (m-RPCA), which is an approach based on distances between moments computed from a robust subspace learned by Robust Principal Component Analysis (RPCA) from contaminated observations, in order to detect anomalies from skewed data and network traffic.

The proposed approach relies on a robust subspace computed from supposed legitimate observations, for estimating the moments to be used for distance analysis. The anomaly detection from contaminated observations evaluate the Mahalanobis distance between the robust moments and new contaminated observations, in a semi-supervised fashion, without the computational cost of new robust subspace learning for anomaly detection from new observations. The m-RPCA can also be computed as an unsupervised algorithm, with subspace learning from the same contaminated data that is the target of the anomaly detection analysis.

We evaluate the accuracy of the m-RPCA for anomaly detection on simulated data sets, with skewed and heavy-tailed distributions, and for network attack detection on CTU-13 data set \cite{garcia2014empirical}, which is a large data set of legitimate, background and botnet traffic that has been adopted to deal with the lack of up-to-date real-world data sets for anomaly detection systems \cite{osanaiye2016distributed}. The Experimental evaluation compares our proposal to standard and widely adopted algorithms for anomaly detection, which are based on clustering and statistical approaches, and to ROBPCA \cite{hubert2005robpca}, which is a anomaly detection method that relies on robust estimates with adjusted outlyingness based on robust skewness.

The main contribution of this work is the proposal of a novel semi-supervised and unsupervised method for anomaly detection in skewed and imbalanced data, with results showing improvements through an experimental evaluation on simulated skewed and heavy-tailed distributions and on real data set with legitimate network traffic and botnet attacks.

This chapter is organized as follows. In Section \ref{sec:3_relatedworks}, a literature review about network anomaly detection, botnet detection, and imbalanced learning is conducted. We present in Section \ref{sec:3_datamodel} the data model and the evaluated data set. In Section \ref{sec:3_proposal} it is described the proposed approach for network attack detection. We discuss in Section \ref{sec:3_experiments} the experimental validation, and in Section \ref{sec:3_simulated_result} are presented the results of the simulated scenarios and in Section \ref{sec:3_ctu13_result} we describe the results for anomaly detection on CTU-13 data set. Finally, in Section \ref{sec:3_conclusion} we draw the conclusions and the suggestions for future work.

\section{Related Works}
\label{sec:3_relatedworks}

@@

According to \cite{distilnetworks2019}, 37.9\% of all Internet traffic of 2018 was not from human activities, but from bots, that can be classified as good bots, that perform legitimate operations, and bad bots, which are responsible by malicious activities, such as DDoS, probe attack or frauds.

@@

Khattak \emph{et al.} \cite{khattak2015botflex} proposed BotFlex, which is a network-based tool for botnet detection, composed by a Complex Event Processing (CEP) engine and on a correlation framework that continuously receives events and correlates them according to rules. BotFlex's results are compared to BotHunter \cite{gu2007bothunter}, but the evaluation relies in an own and not public data set.

@@

Considering the lack of available labeled data sets, Garcia \emph{et al.} \cite{garcia2014empirical} proposes the CTU-13 data set, which is composed by attack, legitimate and background labeled data, in an imbalanced distribution like in a real networks. The authors recommend scenarios for training and testing in order to avoid the use of traffic from a botnet family for training and testing, aiming to ensure that the evaluated methods can generalize and detect new behaviors. Taking into account the adoption of unsupervised or semi-supervised approaches for anomaly detection, adopting the training and testing approach proposed by Garcia \emph{et al.}, some botnet malwares wouldn't be tested, since in the author's proposal some botnets are present only for training.

@@

% Anomaly detection by subspace learning
Traditional PCA-based anomaly detection models are not suitable for anomaly interpretation, as they judge whether a data instance is an anomaly or not based on the length of its projection on the abnormal subspace spanned by the less significant principal components, and there is no direct mapping between PCA’s dimensionality-reduced subspace and the original feature space \cite{ringberg2007sensitivity}. However, to overcome the above mentioned limitations, some approaches based on PCA have been proposed for network anomaly detection.

Callegari \emph{et al.} \cite{callegari2011novel} proposed a PCA-based method for identifying the network traffic flows responsible for an anomaly detected at the aggregate level, by means of a separation of legitimate and anomaly observations according to principal components (legitimate) and remaining (anomalies). Lee \emph{et al.} \cite{Lee2013} presented OverSampling PCA (osPCA), which allows one to determine the anomaly of the target instance according to the variation of the resulting dominant eigenvector obtained by similarity analysis and over sampling. Vieira \emph{et al.} \citeme{vieira2017model} proposed a framework that applies Model Order Selection (MOS) for detection of time frames under attack and uses similarity analysis to extract details and detect the time and ports under attack.

% Anomaly detection by robust subspace learning
The problem of PCA or subspace learning for outlier corrupted data is called Robust Principal Component Analysis (RPCA) or robust subspace learning \cite{candes2011robust, vaswani2018robust}. RPCA aims to be resilient to outliers by means of a robust subspace learning \cite{vaswani2018robust} for outlier corrupted data, decomposing a given data matrix $\pmb{X}$ into the sum of a low rank matrix $\pmb{L}$, whose column subspace gives the principal components, and a sparse matrix $\pmb{S}$, which refers to outliers’ matrix. We refer to \cite{lerman2018overview} and \cite{vaswani2018robust} for more details regarding robust subspace learning.

RPCA has been mainly applied to computer vision, in problems of robust subspace tracking and robust subspace recovery. However, RPCA has also been adopted for general outlier detection \cite{hubert2005robpca,hubert2009robustskewed,cherapanamjeri2017thresholding,zhou2017anomaly,NetflixSurus} and for anomaly detection on network traffic \cite{pascoal2012robust}. ROBPCA \cite{hubert2005robpca} intends to identify outliers using PCA from robust estimates of mean and covariance matrix, to reduce the data dimensions and plotting the orthogonal distances versus the robust score distances, to flag an outlier map. However, ROBPCA flags many points as outlying when the original data is skewed. Therefore, Hubert \emph{et al.} \cite{hubert2009robustskewed} proposed ROBPCA-AO, which improves ROBPCA for problems with skewed data, by means of an adjusted outlyingness based on robust skewness. Hubert \emph{et al.} \cite{hubert2009robustskewed} evaluated ROBPCA-AO for real and simulated data sets, but it is not clear if this method was evaluated for very skewed and imbalanced data sets, such as in the network attack detection problem.

Robust subspace learning has received a growing attention of researchers aiming the development of network anomaly detection systems \cite{ hubert2009robustskewed, pascoal2012robust, zhou2017anomaly}, considering outlier-robust methods and sparse-corruption methods \cite{lerman2018overview}. Pascoal \emph{et al.} \cite{pascoal2012robust} proposed an approach based on a robust mutual information estimator for feature selection and based on RPCA for outlier detection in internet traffic. The anomaly detection proposed by Pascoal \emph{et al.} is an unsupervised approach that estimate the first $k$ robust principal components, calculate the score and the orthogonal distances, calculate the thresholds and classify new observations accordingly.

Zhou and Paffenroth \cite{zhou2017anomaly} proposed the Robust Deep Autoencoders (RDA), which the central idea is that a RDA inherits the non-linear representation capabilities of autoencoders combined with the anomaly detection capabilities of RPCA. Considering that outliers and noise may reduce the quality of representations discovered by deep autoencoders, the proposed model isolates noise and outliers in the input by means of a RPCA approach, and the autoencoder is trained after this isolation. RDA was evaluated by the authors for the MNIST data set.

@@

Mahalanobis Distance (MD) is a generalized distance which is useful for determining the similarity between an unknown sample and a collection of known samples, by considering the covariance between the variables and their mean values. The MD is a measure of the distance between a vector $\pmb{x}$ and a distribution $\pmb{X}$, introduced by P. C. Mahalanobis in 1936 \cite{mahalanobis1936md}. MD is a multi-dimensional generalization for measuring how many standard deviations away $\pmb{x}$ is from the mean $\pmb{\mu}$ and covariance $\hat{\pmb{\Sigma}}$ of $\pmb{X}$. MD has been used for distance based anomaly detection with robust estimates in many areas, assuming that the Mahalanobis Distance between robust estimates and contaminated observations can reveal anomalies.

We propose an approach based on the distances between moments computed from learned robust subspace and contaminated observations, for anomaly detection on skewed and imbalanced data sets, and evaluate our approach for network attack detection. The proposed approach relies on learning robust subspace from supposed legitimate traffic for estimating the moments (mean, skewness and kurtosis).

Thus, the anomaly detection for contaminated observations should evaluate the distance between the robust moments and contaminated observations, in a semi-supervised fashion and without the computational cost of new robust subspace learning for new observations, or in an unsupervised approach, where the robust subspace is learned from the contaminated data and the distance analysis is between the robust moments and the contaminated observations.

\section{Data Model}
\label{sec:3_datamodel}

In this chapter, we adopt the data model introduced in Section \ref{sec:2_datamodel} and define that the Frobenius norm is denoted as $\left\| \mathord{\cdot} \right\|_F$, while $\left\| \mathord{\cdot} \right\|_*$ denotes the nuclear norm of a matrix and $\left\| \mathord{\cdot} \right\|_1$ means the sum of the absolute values of matrix entries. We also define the operator $\langle \mathord{\cdot} \rangle$, that is the standard trace inner product, and the operator $[ \mathord{\cdot} ]^c$, which denotes the indexes of the $c$ largest values of a vector.

This section also presents a description of the simulated data model in Subsection \ref{sec:3_simulation} and in Subsection \ref{sec:3_CTU-13} we describe the data model for the CTU-13 data set.

\subsection{The Simulated data set}
\label{sec:3_simulation}

In order to analyze the hypotheses $H_2^{(N)}$ and $H_2^{(A)}$, which evaluate if the robust subspace learning can improves the anomaly detection from imbalanced and skewed data, we create two simulated data sets characterized by skewed and heavy tailed distributions, and create a simulated Gaussian distributed data set, in order to also analyze the detection rate for not skewed and not heavy tailed distributions.

We selected Pareto (with scale $a=3$, mode $m=1$ and $\mu=1$) and Lognormal (with mean $\mu=0$ and standard deviation $\sigma=1$) distributions, denoted respectively as $\pmb{Y}_p$, while $\pmb{Y}_l$, to simulate legitimate events and to evaluate the anomaly detection on skewed data performed by our proposals in comparison to widely adopted algorithms for outlier detection, considering that these distributions are skewed and heavy-tailed, and have been adopted to characterize network traffic of the Internet and data centers \cite{benson2010network,leon2017probability}. We also adopt the Gaussian distributed data set $\pmb{Y}_g$ to simulate legitimate events with mean $\mu=0$ and unitary standard deviation $\sigma=1$.

@@

An uniform noise between $-6$ and $6$ is used to add anomalies into the Gaussian distributed data set, and $\pmb{Y}_g^c$ denotes the contaminated Gaussian data contaminated by an uniform distribution. The figure \ref{fig:4.04} shows an example of the Gaussian distribution of legitimate observations and the uniform distribution to simulate the addition of anomalies.

@@

\subsection{The CTU-13 data set}
\label{sec:3_CTU-13}

The CTU-13 \cite{garcia2014empirical} is a data set of botnet traffic that was captured in the Czech Technical University, by means of a testbed and malware execution in a real network. The CTU-13 data set contains 13 scenarios with network flows of botnet malwares, that are: neris, rbot, virut, menti, sogou, nsys.ay and murlo. The botnet traffic is also classified as attack or command and control (C\&C), while the legitimate flows can also be classified as legitimate or background.

@@

\begin{itemize}
	\item \pmb{Attacks:} Click Fraud (CF), Port Scan (PS), Fast Flux (FF), SPAM and DDoS;
	\item \pmb{C\&C:} IRC, P2P and HTTP.
\end{itemize}

We refer to Garcia \cite{garcia2014identifying} and Garcia \emph{et al.} \cite{garcia2014empirical} for a detailed description of the performed attacks and C\&C flows, as well as for more information about the topology of the adopted testbed, rules for classifying legitimate flows, and an analysis of behaviors or patterns of the malware's traffic.

For all the scenarios, the authors of the CTU-13 data set convert the captured pcap files to NetFlows and release the processed flows. The data set provides ground-truth labels for flows as follows: flows from or to the infected machines are labeled as “botnet”; flows from or to well-known and controlled machines are labeled as “normal”; all other flows are labeled as “background.”

Table \ref{tab:4.01} presents an overview grouped by scenario, according to the column ID, and shows the malwares used for botnet attacks, the types of attacks or C\&C, the total number of flows, the number of malicious flows which includes flows of C\&C and attacks, and finally shows the number of legitimate flows.

\begin{table}[h!]
	\scriptsize
	\caption{CTU-13 data set description}
	\label{tab:4.01}
	\begin{tabular}{| l | l | l | r | r | r | r | r | r | r | r | }
		\hline \rowcolor{Gray} \begin{tabular}[x]{@{}l@{}}\textbf{ID}\end{tabular}	&\begin{tabular}[x]{@{}l@{}}\textbf{Bot}\end{tabular}	 &\begin{tabular}[x]{@{}l@{}}\textbf{Type}\end{tabular}	&\begin{tabular}[x]{@{}l@{}}\textbf{Total}\end{tabular} &\begin{tabular}[x]{@{}l@{}}\textbf{Malicious}\end{tabular} &\begin{tabular}[x]{@{}l@{}}\textbf{C\&C}\end{tabular} &\begin{tabular}[x]{@{}l@{}}\textbf{Attack}\end{tabular} &\begin{tabular}[x]{@{}l@{}}\textbf{Normal}\end{tabular}\\ \hline
			10 & neris &\makecell[l]{IRC, Spam,\\CF} & 2,824,636 & 40,961 (1.45\%) & 341 (0.01\%) & 40,620 (1.44\%) &30,387 (1.07\%)\\ \hline
			11 & neris &\makecell[l]{IRC, Spam,\\CF} & 1,808,122 & 20,941 (1.16\%) & 673 (0.04\%) & 20,268 (1.12\%) &9,120 (0.5\%)\\ \hline
			12 & rbot &\makecell[l]{IRC, PS} & 4,710,638 & 26,822 (0.57\%) & 63 (0.00\%) & 26,759 (0.57\%) &116,887 (2.48\%)\\ \hline
			15 & rbot &\makecell[l]{IRC, DDoS} & 1,121,076 & 2,580 (0.23\%) & 52 (0.00\%) & 2,528 (0.23\%) &25,268 (2.25\%)\\ \hline
			15-2 & virut &\makecell[l]{Spam, PS,\\HTTP} & 129,832 & 901 (0.69\%) & 24 (0.02\%) & 877 (0.68\%) &4,679 (3.6\%)\\ \hline
			16 & menti &PS & 558,919 & 4,630 (0.83\%) & 199 (0.04\%) & 4,431 (0.79\%) &7,494 (1.34\%)\\ \hline
			16-2 & sogou &HTTP & 114,077 & 63 (0.06\%) & 26 (0.02\%) & 37 (0.03\%) &1,677 (1.47\%)\\ \hline
			16-3 & murlo &PS & 2,954,230 & 6,127 (0.21\%) & 1,074 (0.04\%) & 5,053 (0.17\%) &72,822 (2.46\%)\\ \hline
			17 & neris &\makecell[l]{IRC, Spam,\\CF, PS} & 2,087,508 & 184,987 (8.86\%) & 2,973 (0.14\%) & 182,014 (8.72\%) &43,340 (1.57\%)\\ \hline
			18 & rbot &\makecell[l]{IRC, DDoS} & 1,309,791 & 106,352 (8.12\%) & 33 (0.00\%) & 106,319 (8.12\%) &15,847 (1.2\%)\\ \hline
			18-2 & rbot &\makecell[l]{IRC, DDoS} & 107,251 & 8,164 (7.61\%) & 2 (0.00\%) & 8,162 (7.61\%) &2,718 (2.53\%)\\ \hline
			19 & nsys.ay &P2P & 325,471 & 2,168 (0.67\%) & 25 (0.01\%) & 2,143 (0.66\%) &7,628 (2.35\%)\\ \hline
			15-3 & virut &\makecell[l]{Spam, PS,\\HTTP} & 1,925,149 & 40,003 (2.08\%) & 536 (0.03\%) & 39,467 (2.05\%) &31,939 (1.65)\\ \hline
	\end{tabular}
\end{table}

The full data set and scenarios can be denoted as $\pmb{X} = \{\pmb{X}_{10}, \pmb{X}_{11}, \ldots , \pmb{X}_{18-2}, \pmb{X}_{19}\}$, in accordance to IDs presented in Table \ref{tab:4.01}. In our experiment each contaminated scenario $\pmb{X}_i$ is split into $\pmb{X}_i^s$ containing 50\% of the legitimate data, and into $\pmb{X}_i^c$ that is composed by all anomalous flows and the necessary number of legitimate flows to have a testing data with the desired contamination rate.
